---
title: "Lab1"
author: "KatieLawler"
date: "`r Sys.Date()`"
output: html_document
---

#Setting Up Packages

```{r}
install.packages("pacman") # command return to run a line of code
pacman::p_load(haven, readr, dplyr, openxlsx)
```

**#1. Import class one survey data from our Github site (<https://github.com/kijohnson/Advanced-Data-Analysis>) (The dataset is called ‘Class 1 Survey Fall 2025 2.csv’), calling the R dataframe that you create C1survey. #1 Answer Below:**

```{r, Answer1, message=FALSE}
lab1_csv <-
 read_csv(
   "https://github.com/kijohnson/Advanced-Data-Analysis/raw/refs/heads/main/Class%201%20Survey%20Fall%202025%202.csv"
   )
```

**#2. Determine: a. the number of observations (i.e. the number of people who filled out the survey) and b. the number of variables in the dataframe.**

#2A Answer: 58 Observations (i.e. rows)

```{r, Answer2, message=FALSE}
dim(lab1_csv) 
```

#2B Answer: 27 Variables (i.e. columns)

```{r, Answer2B, message=FALSE}
dim(lab1_csv) 
```

**#3. a. Rename the column variables to something shorter and that is descriptive of what the variable is about (for example like_dogs for the ‘Do you like dogs?’ question variable) and b. Display that they are renamed. Hints: You can use either the rename_with function or the names function. The list of names for time is:**

#3A Answer:

```{r}
lab1_csv <- lab1_csv %>%
  rename_with(~ c(
    "id", "like_cats", "like_dogs", "have_desert", "slogan", 
    "fav_day", "larkORowl", "fav_food", "fav_drink", "fav_season", 
    "fav_month", "hobby", "program", "specialization", "stat_software", 
    "R_exp", "coding_comfort", "coding_length", "top_three", 
    "public_health_interest", "fav_num", "bday", "bmonth", 
    "country", "state", "city", "highest_educ_level"
  ))
```

#3B Answer:

```{r}
names(lab1_csv)
```

**#4. Write code to determine and display the number of factor, integer, numerical, and character variables there are in the *C1survey* dataset. Suggestion: Try to ask ChatGPT for a solution if you are unsure. There is a really short answer to this but we want to see what students come up with!**

#4 Answer:

```{r}
sapply(lab1_csv, class) |> 
  table()

```

**#5. a. Using code, check the *bday* and *bmonth* variables for any unusual or missing values (hint: use table function). If you find any, b. describe what you will do with the unusual values in a comment before or immediately following the code that does it, and c. after you have cleaned up any unusual values, find the median bday and month. \*Note: please make sure that your bday and month variables are numeric before doing statistics**

#5A Answer:

```{r}
table(lab1_csv$bday, useNA = "ifany")

table(lab1_csv$bmonth, useNA = "ifany")

```

#5B Answer: With any unusual values, I will treat than as NA

```{r}
# Making sure all data is in numeric form

month_map <- c(
  "January" = 1,  "February" = 2,  "March" = 3,     "April" = 4,
  "May" = 5,      "June" = 6,      "July" = 7,      "August" = 8,
  "September" = 9,"October" = 10,  "November" = 11, "December" = 12
)

lab1_csv$bmonth <- ifelse(
  lab1_csv$bmonth %in% names(month_map), 
  month_map[lab1_csv$bmonth],  # replace with corresponding number
  lab1_csv$bmonth              # leave other values as-is (numbers or NA)
)

lab1_csv$bmonth <- as.numeric(trimws(lab1_csv$bmonth))

# Convert UVs to NAs
# Bday valid range = 1–31
lab1_csv$bday[ lab1_csv$bday < 1 | lab1_csv$bday > 31 ] <- NA

# Bmonth valid range = 1–12
lab1_csv$bmonth[ lab1_csv$bmonth < 1 | lab1_csv$bmonth > 12 ] <- NA

```

#5C Answer:

```{r}
# Change to numeric
lab1_csv$bday <- as.numeric(lab1_csv$bday)
lab1_csv$bmonth <- as.numeric(lab1_csv$bmonth)

# Medians (TRUE ignores NAs)
median(lab1_csv$bday, na.rm = TRUE)
median(lab1_csv$bmonth, na.rm = TRUE)

```

**#6. a. Create a new variable called *bseason* that gives the season according to Northern Meteorological season in which respondents were born (winter = 12/1 to 2/29, spring = 3/1 to 5/31, summer = 6/1 to 8/31, fall = 9/1 to 11/30). b. Using the table function, print a table of Seasons in the columns and bmonths in the rows that allows you to check that the coding is correct c. Sum the columns using the addmargins function to answer the question of how many classmates were born in each season? Include your answer in the code.**

#6A Answer:

```{r}
lab1_csv$bseason <- with(lab1_csv,
  ifelse(bmonth %in% c(12, 1, 2), "Winter",
  ifelse(bmonth %in% c(3, 4, 5), "Spring",
  ifelse(bmonth %in% c(6, 7, 8), "Summer",
  ifelse(bmonth %in% c(9, 10, 11), "Fall", NA)))))
```

#6B Answer:

```{r}

season_table <- table(lab1_csv$bmonth, lab1_csv$bseason)

print(season_table)

```

#6C Answer:

```{r}
season_table_with_sums <- addmargins(season_table)

print(season_table_with_sums)
```

### 

**7. Pick your favorite variable or variables to analyze, come up with a question you want to answer with that variable or variables, generate the code, and provide an answer to your question. Describe what you did using comments (i.e. #’s).**
