---
title: "Lab1 Analysis Code"
author: "KatieLawler"
date: "`r Sys.Date()`"
output: html_document
---

**#Setting Up Packages**

```{r}
install.packages("pacman") # command return to run a line of code
pacman::p_load(haven, readr, dplyr, openxlsx)
```

**#1. Import class one survey data from our Github site (<https://github.com/kijohnson/Advanced-Data-Analysis>) (The dataset is called ‘Class 1 Survey Fall 2025 2.csv’), calling the R dataframe that you create C1survey. #1 Answer Below:**

```{r, Answer1, message=FALSE}
lab1_csv <-
 read_csv(
   "https://github.com/kijohnson/Advanced-Data-Analysis/raw/refs/heads/main/Class%201%20Survey%20Fall%202025%202.csv"
   )

```

**#2. Determine: a. the number of observations (i.e. the number of people who filled out the survey) and b. the number of variables in the dataframe.**

#2A Answer: 58 Observations (i.e. rows)

```{r, Answer2, message=FALSE}
dim(lab1_csv)  #look at dimensions of the dataframe

```

#2B Answer: 27 Variables (i.e. columns)

```{r, Answer2B, message=FALSE}
dim(lab1_csv) #look at dimensions of the dataframe

```

**#3. a. Rename the column variables to something shorter and that is descriptive of what the variable is about (for example like_dogs for the ‘Do you like dogs?’ question variable) and b. Display that they are renamed. Hints: You can use either the rename_with function or the names function. The list of names for time is:**

#3A Answer:

```{r}
lab1_csv <- lab1_csv %>%
  rename_with(~ c(
    "id", "like_cats", "like_dogs", "have_desert", "slogan", 
    "fav_day", "larkORowl", "fav_food", "fav_drink", "fav_season", 
    "fav_month", "hobby", "program", "specialization", "stat_software", 
    "R_exp", "coding_comfort", "coding_length", "top_three", 
    "public_health_interest", "fav_num", "bday", "bmonth", 
    "country", "state", "city", "highest_educ_level"
  ))  #use C for concatenate for whole list to be changed

```

#3B Answer:

```{r}
names(lab1_csv) #check

```

**#4. Write code to determine and display the number of factor, integer, numerical, and character variables there are in the *C1survey* dataset. Suggestion: Try to ask ChatGPT for a solution if you are unsure. There is a really short answer to this but we want to see what students come up with!**

#4 Answer:

```{r}
sapply(lab1_csv, class) |> 
  table() #use apply and table functions

```

**#5. a. Using code, check the *bday* and *bmonth* variables for any unusual or missing values (hint: use table function). If you find any, b. describe what you will do with the unusual values in a comment before or immediately following the code that does it, and c. after you have cleaned up any unusual values, find the median bday and month. \*Note: please make sure that your bday and month variables are numeric before doing statistics**

#5A Answer:

```{r}
table(lab1_csv$bday, useNA = "ifany") #use $ for specific column, useNA function

table(lab1_csv$bmonth, useNA = "ifany") #use $ for specific column, useNA function

```

#5B Answer: With any unusual values, I will convert them to the proper number format, if they do not fit in or make sense (ex. birth date is 58) or are blank I will then treat as Not Available, or NA.

```{r}
# Making sure all data is in numeric form (also accounting for mispelling, ex. see september)

month_map <- c(
  "January" = 1, "February" = 2, "March" = 3, "April" = 4,
  "May" = 5, "June" = 6, "june" = 6, "July" = 7, "August" = 8,
  "September" = 9, "Septemeber" = 9, "October" = 10,
  "November" = 11, "December" = 12
)



lab1_csv$bmonth <- ifelse(
  lab1_csv$bmonth %in% names(month_map), 
  month_map[lab1_csv$bmonth],  # Replace with number for that month
  lab1_csv$bmonth              # Leave other values as was found (can clear empty/NA later)
)

lab1_csv$bmonth <- as.numeric(trimws(lab1_csv$bmonth))

# Convert UVs to NAs since bday valid range = 1–31
lab1_csv$bday[ lab1_csv$bday < 1 | lab1_csv$bday > 31 ] <- NA

# Do the same here, but bmonth valid range = 1–12
lab1_csv$bmonth[ lab1_csv$bmonth < 1 | lab1_csv$bmonth > 12 ] <- NA

```

#5C Answer:

```{r}
# Change to Numeric
lab1_csv$bday <- as.numeric(lab1_csv$bday)
lab1_csv$bmonth <- as.numeric(lab1_csv$bmonth)

# Calculating Class Medians (setting =TRUE ignores NAs when counting)
median(lab1_csv$bday, na.rm = TRUE)
median(lab1_csv$bmonth, na.rm = TRUE)

```

**#6. a. Create a new variable called *bseason* that gives the season according to Northern Meteorological season in which respondents were born (winter = 12/1 to 2/29, spring = 3/1 to 5/31, summer = 6/1 to 8/31, fall = 9/1 to 11/30). b. Using the table function, print a table of Seasons in the columns and bmonths in the rows that allows you to check that the coding is correct c. Sum the columns using the addmargins function to answer the question of how many classmates were born in each season? Include your answer in the code.**

#6A Answer:

```{r}
lab1_csv$bseason <- with(lab1_csv,
  ifelse(bmonth %in% c(12, 1, 2), "Winter",
  ifelse(bmonth %in% c(3, 4, 5), "Spring",
  ifelse(bmonth %in% c(6, 7, 8), "Summer",
  ifelse(bmonth %in% c(9, 10, 11), "Fall", NA)))))
```

#6B Answer:

```{r}

season_table <- table(lab1_csv$bmonth, lab1_csv$bseason)

print(season_table) #See counts for each season

```

#6C Answer:

```{r}
season_table_with_sums <- addmargins(season_table)

print(season_table_with_sums)
```

### 

**7. Pick your favorite variable or variables to analyze, come up with a question you want to answer with that variable or variables, generate the code, and provide an answer to your question. Describe what you did using comments (i.e. #’s).**

Question: What is our favorite number as an ADA class?

```{r}
#Count up how many times each number appears

fav_num_counts <- table(lab1_csv$fav_num)
print(fav_num_counts)

#For output, top numbers are inputs from class, bottoms are the counts


```

```{r}

#Number with highest count is our class' favorite

#So the max number count is class favorite

favorite_number <- as.numeric(names(fav_num_counts)[which.max(fav_num_counts)])
cat("The class's favorite number is:", favorite_number, "\n")
```
